---
title: "Project1"
format: html
editor: visual
---

## Annie DiFrank

```{r setup, include=FALSE}
#this chunk is just to install/load needed packages 
knitr::opts_chunk$set(echo = TRUE)

# Use pacman to check whether packages are installed, if not load them
if (!require("pacman")) install.packages("pacman")
library(pacman)
pacman::p_load(tidycensus,
 dplyr,
 tidyverse,
 httr,
 jsonlite)

```

# Part 1: Data Processing

### In this section, we will query the API and return a nicely formatted tibble.

Process:

-   Build the appropriate URL

-   Use httr:get() to contact the website

-   Write a function to parse the data and put it into a tibble for ease

```{r}
#Build the URL 
pums_url<- "api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24"
#Get the URL data by contacting the website 
url_info<- httr::GET(pums_url)
str(url_info,max.level=1)

#Helper function to take output from GET() and turn it into a tibble 
get_tibble<- function(url_info) {
  #Parse the content
  parsed<- fromJSON(rawToChar(url_info$content))
  tibble_data<- as_tibble(parsed)
  
  #set the first row to the names of the columns 
  names(tibble_data) <- as.character(unlist(tibble_data[1,]))
  #remove the old first row 
  tibble_data<-tibble_data[-1,]
  
  return(tibble_data)
}
#testing function 
test<- get_tibble(url_info = url_info)

```

### In this section, we will write a function that allows users to query the API and change modifiers

```{r eval=FALSE}
query_pums <- function(
  year = 2022,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = c("SEX"),
  geography_level = "All",
  geography_filter_value = NULL
) {
  # Validate the year
  if (!(year %in% 2010:2022)) {
    stop("Year must be between 2010 and 2022")
  }
  
    # defining the numeric and categorical variable options
    num_var_options <-
      c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
    cat_var_options <-
      c("FER",
        "HHL",
        "HISPEED",
        "JWAP",
        "JWDP",
        "JWTRNS",
        "SCH",
        "SCHL",
        "SEX")
    
    # making sure the variables asked for are in this set of variables 
      if (any(!numeric_vars %in% num_var_options)) {
    stop("Invalid numeric variables provided")
      }
    
    # making sure one numeric variable other than PWGTP is included
if (length(numeric_vars) < 1) {
    stop("At least one numeric variable other than PWGTP must be provided")
}
        
   # making sure the variables asked for are in this set of variables 
    if (any(!categorical_vars %in% cat_var_options)) {
    stop("Invalid categorical variables provided")
    }
  
    # making sure at least one categorical variable is provided
  if (length(categorical_vars) < 1) {
    stop("At least one categorical variable must be provided")
  } 
  
  # specifying the geographic level
  valid_geo_levels <- c("All", "Region", "Division", "State")
  if (!(geography_level %in% valid_geo_levels)) {
    stop("Geography level is invalid")
  }
  
  # making the API request- setting the base of the url
  base_url <- "https://api.census.gov/data/"
  api_url <- paste0(base_url,as.character(year), "/acs/acs1/pums?get=")
  
  # Construct API parameters
  params <- list(
    get = paste(c(numeric_vars, categorical_vars), collapse = ",")
  )
  
  # Setting geography parameter
  if (geography_level == "All") {
    params$geo_level <- NULL
  } else if (geography_level == "Region") {
    params$geo_level <- "region:*"
  } else if (geography_level == "Division") {
    params$geo_level <- "division:*"
  } else if (geography_level == "State") {
    params$geo_level <- "state:*"
  }
  
    # Add geography filter if specified
  if (!is.null(geography_filter_value)) {
    params$geo_level <- geography_filter_value
  }
  
  # Print params for debugging
  print("API URL:")
  print(api_url)
  print("API parameters:")
  print(params)
  
  #appending parameters to the api url 
  api_url<- paste0(api_url, params$get)
  #optional geography level 
  if (!is.null(params$geo_level)) { 
    #append geo param 
    api_url<- paste0(api_url, "&for=", params$geo_level)}
  
  # Query the API
  response <- GET(api_url)
  # Parse response
  data <- content(response, as = "text") %>%
    fromJSON(flatten = TRUE) %>%
    as_tibble()
  
    #set the first row to the names of the columns 
  names(data) <- as.character(unlist(data[1,]))
  #remove the old first row 
  data<-data[-1,]
  
      # convert numeric variables to numeric values
    num_vars_tibble <- intersect(numeric_vars, names(data))
    data[num_vars_tibble] <- lapply(data[num_vars_tibble], as.numeric)
    
    # handle time variables, convert to middle of time period 
    time_variables <- c("JWAP", "JWDP", "JWMNP")
    time_vars_tibble <- intersect(time_variables, names(data))
    data[time_vars_tibble] <-
      lapply(data[time_vars_tibble], function(x) {
        as.numeric(x) / 2
      })
  
  # convert categorical variables to factors
  cat_vars_tibble <- intersect(categorical_vars, names(data))
  data[cat_vars_tibble] <- lapply(data[cat_vars_tibble], as.factor)
  
  #adding a class of 'census' to the tibble 
  class(data) <- c("census", class(your_tibble)
  
  return(data)
}


#testing this 
test2 <- query_pums(
  year = 2021,
  numeric_vars = c("AGEP"),
  categorical_vars = c("SEX"),
  geography_level = "All")

# year=2021
# numeric_vars = c("AGEP")
#  categorical_vars = c("SEX")
#   geography_level = "All"

```

### Next, we will write a function that allows the user to specify multiple years of survey data (and all the other options above)

# Part 2: Writing a summarizing function

### Here, we will process the data in the tibble appropriately.

-   Going to write a generic function for summarizing, which first involves making the tibble created in part 1 have an additional class of 'census'

-   Next, will write a custom summary function that gives us the means and std devs for our numeric variables and counts for our categorical variables

-   Additionally, will create a generic plot() function for a census class table

```{r eval = FALSE}
#background- analyzing plot functions
plot.function #what is used for a class=function
getS3method("plot","data.frame") #what is used for a class=data frame 
plot #generic method that specifies UseMethod("plot)

#First, we will update the above function to create the API data and make the tibble have an additional class of 'census'. Rather than repeating the lengthy function in this chunk, I have appended it to the original function

#Next, writing a custom summary function for class = census 

summary.census <- function(tibble, numeric_variables, categorical_variables) {}
```
